A performance comparison of Dask and Apache Spark for data-intensive neuroimaging pipelines

### 2019-07-30 latex/paper.tex
fixed minor typos in abstract to methods

In the past few years, neuroimaging has entered the Big Data era due to the joint
increase in image resolution, data sharing, and study sizes. However, no particular
Big Data engines has emerged in this field, and several alternatives remain
available. We compare two popular Big Data engines with Python APIs, Apache Spark and
Dask, for their runtime performance in processing neuroimaging pipelines. Our
evaluation uses two synthetic pipelines processing the \SI{81}{\giga\byte} Big Brain
image, and a real pipeline processing anatomical data from more than 1,000 subjects.
We benchmark these pipelines using various combinations of task durations, data
sizes, and numbers of workers, deployed on a 8-node (8 cores ea.) compute cluster in
Compute Canada's Arbutus cloud. We evaluate PySpark's RDD API against Dask's Bag,
Delayed and Futures. Results show that despite slight differences between Spark and
Dask, both engines perform comparably. However, Dask pipelines risk being limited
by Python's GIL depending on task type and cluster configuration. In all cases, the
major limiting factor was data transfer. While either engine is suitable for
neuroimaging pipelines, more effort needs to be placed in reducing data transfer time.
\end{abstract}

@@

The next section details the design of our benchmarking experiments. We consider two
data-intensive neuroimaging applications: high-resolution imaging,
represented by the Big Brain data~\cite{Amunts:13}, and large functional
MRI studies, represented by data from the consortium for reliability and
reproducibility (CoRR~\cite{zuo2014open}). We test application pipelines
involving different patterns (map-only, map-reduce), and different types of
implementations (plain Python, command-line,
containerized). We evaluate performance on a dedicated cluster
representative of the ones used in today's data-intensive neuroimaging
studies, using the main Dask and Spark APIs. The other sections present
our results, discussion and conclusions.

@@

\subsubsection{Apache Spark} Apache Spark is a widely-used 
general-purpose Big Data engine.
Its main abstraction, the Resilient Distributed Dataset
(RDD)~\cite{RDD}, is a fault-tolerant, parallel collection of data elements. It 
achieves fault tolerance through the recording of data lineage, the
sequence of operations used to modify the original data. The RDD is the basis
of Spark's other data structures, namely, Datasets
and DataFrames. Datasets are similar to RDDs but additionally use Spark SQL's
optimized execution engine to further improve performance. DataFrames, used to process
tabular data, are Datasets where the data is organized into named-columns.
While the DataFrame API exists in all the available language APIs, Datasets are
limited to Scala and Java.

@@

% Didn't mention Kubernetes as it's still experimental
Spark is compatible with three different schedulers: Spark
standalone, YARN and Mesos.
The Spark standalone scheduler is a simple default scheduler built into Spark.
In contrast, the YARN scheduler is primarily designed to schedule Hadoop-based 
workflows, whereas Mesos can be used to schedule
a variety of different workflows. As researchers would likely be
executing their workflows in HPC environments with neither YARN or Mesos
installed, we limit our focus to Spark's Standalone scheduler.

@@

\subsubsection{Dask} Dask is a Python-based Big Data engine that is becoming
increasingly popular in the scientific Python ecosystem. Like Spark, Dask avoids data transfers and needless computation and communication through
in-memory computing, data locality, and lazy evaluation. Dask worklflows can 
further reduce data transfer costs by leveraging multithreading whenever 
communication is not bounded by Python's GIL. Dask relies on data 
lineage to achieve fault tolerance, however, unlike Spark, does not require
operations to be coarse grained. Furthermore, Dask is lightweight and modular,
allowing users to only install the components they require.

@@

Three neuroimaging applications to evaluate the engines in different
conditions were used. The first two, incrementation and histogram, are simple 
synthetic applications representing map-only and map-reduce applications,
respectively. The third one is a real neuroimaging application representative
of popular BIDS applications~\cite{gorgolewski2017bids}. All scripts used for our
experiments are available at
\url{https://github.com/big-data-lab-team/paper-big-data-engines}

@@

The number of data partitions also has an impact on application
performance. In our experiments, we set a partition for each block or
subject, resulting in a maximum of 750 partitions for the Big Brain
applications, and 1,397 partitions for the BIDS app. The number of
partitions obeys a classical trade-off: small numbers of partitions reduce 
overhead, while large numbers of
partitions increase parallelism. Since both Spark and Dask can
work with custom partition sizes, we do not expect substantial differences
coming from variations of this parameter.

>>>>

In the past few years, neuroimaging has entered the Big Data era due to the joint
increase in image resolution, data sharing, and study sizes. However, no particular
Big Data engines has emerged in this field, and several alternatives remain
available. We compare two popular Big Data engines with Python APIs, Apache Spark and
Dask, for their runtime performance in processing neuroimaging pipelines. Our
evaluation uses two synthetic pipelines processing the \SI{81}{\giga\byte} BigBrain
image, and a real pipeline processing anatomical data from more than 1,000 subjects.
We benchmark these pipelines using various combinations of task durations, data
sizes, and numbers of workers, deployed on a 8-node (8 cores ea.) compute cluster in
Compute Canada's Arbutus cloud. We evaluate PySpark's RDD API against Dask's Bag,
Delayed and Futures. Results show that despite slight differences between Spark and
Dask, both engines perform comparably. However, Dask pipelines risk being limited
by Python's GIL depending on task type and cluster configuration. In all cases, the
major limiting factor was data transfer. While either engine is suitable for
neuroimaging pipelines, more effort needs to be placed in reducing data transfer time.
\end{abstract}

@@

The next section details the design of our benchmarking experiments. We consider two
data-intensive neuroimaging applications: high-resolution imaging,
represented by the BigBrain data~\cite{Amunts:13}, and large functional
MRI studies, represented by data from the consortium for reliability and
reproducibility (CoRR~\cite{zuo2014open}). We test application pipelines
involving different patterns (map-only, map-reduce), and different types of
implementations (plain Python, command-line,
containerized). We evaluate performance on a dedicated cluster
representative of the ones used in today's data-intensive neuroimaging
studies, using the main Dask and Spark APIs. The other sections present
our results, discussion and conclusions.

@@

\subsubsection{Apache Spark} Apache Spark is a widely-used 
general-purpose Big Data engine.
Its main abstraction, the Resilient Distributed Dataset
(RDD)~\cite{RDD}, is a fault-tolerant, parallel collection of data elements. It 
achieves fault tolerance through the recording of data lineage, the
sequence of operations used to modify the original data. The RDD is the basis
of Spark's other data structures, namely, Datasets
and DataFrames. Datasets are similar to RDDs, but additionally use Spark SQL's
optimized execution engine to further improve performance. DataFrames, used to process
tabular data, are Datasets where the data is organized into named-columns.
While the DataFrame API exists in all the available language APIs, Datasets are
limited to Scala and Java.

@@

% Didn't mention Kubernetes as it's still experimental
Spark is compatible with three different schedulers: Spark
Standalone, YARN and Mesos.
The Spark Standalone scheduler is a simple default scheduler built into Spark.
In contrast, the YARN scheduler is primarily designed to schedule Hadoop-based 
workflows, whereas Mesos can be used to schedule
a variety of different workflows. As researchers would likely be
executing their workflows in HPC environments with neither YARN or Mesos
installed, we limit our focus to Spark's Standalone scheduler.

@@

\subsubsection{Dask} Dask is a Python-based Big Data engine that is becoming
increasingly popular in the scientific Python ecosystem. Like Spark, Dask avoids data transfers and needless computation and communication through
in-memory computing, data locality, and lazy evaluation. Dask workflows can 
further reduce data transfer costs by leveraging multithreading whenever 
communication is not bounded by Python's GIL. Dask relies on data 
lineage to achieve fault tolerance, however, unlike Spark, does not require
operations to be coarse grained. Furthermore, Dask is lightweight and modular,
allowing users to only install the components they require.

@@

We used three neuroimaging applications to evaluate the engines in different
conditions. The first two, incrementation and histogram, are simple 
synthetic applications representing map-only and map-reduce applications,
respectively. The third, is a real neuroimaging application representative
of popular BIDS applications~\cite{gorgolewski2017bids}. All scripts used for our
experiments are available at
\url{https://github.com/big-data-lab-team/paper-big-data-engines}

@@

The number of data partitions also has an impact on application
performance. In our experiments, we set a partition for each block or
subject, resulting in a maximum of 750 partitions for the BigBrain
applications, and 1,397 partitions for the BIDS app. The number of
partitions obeys a classical trade-off: small numbers of partitions reduce 
overhead, while large numbers of
partitions increase parallelism. Since both Spark and Dask can
work with custom partition sizes, we do not expect substantial differences
coming from variations of this parameter.
